---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r} 
datasets <- tt_load("2020-09-22")
names(datasets)
expeditions <- datasets$expeditions
view(expeditions)
?fct_lump()

lumped_peaks <- expeditions %>%
  mutate(peak_name = forcats::fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != "Other")

view(lumped_peaks)
str(lumped_peaks)
?scale_fill_viridis_c

#please click on "Zoom" to see in real size, now the sizes are a bit messed up with the small window. 
ggplot(lumped_peaks, aes(y= forcats::fct_reorder(peak_name, 
                                                 .x=peak_name,
                                                 .fun = length),
                                                 fill= season))+
  scale_fill_viridis_d()+
  geom_bar()+
  labs(title="The 15 most popular peaks stacked by season of expedition",
       x="Number of expeditions",
       y= NULL)+
  theme(plot.title=element_text(hjust=0.5, 
                                size= 25),
        axis.title.x= element_text(size=16),
        axis.text.y = element_text(size = 10),
        legend.position= "bottom",
        panel.background = element_rect(fill= NA),
        panel.grid= element_line(color= "grey"),
        panel.border = element_rect(color= "grey", fill = NA)
        )




```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
library(scales)
data_task2 <- tt_load("2019-02-19")
names(data_task2)
phds <- data_task2$phd_by_field
view(phds)



agg_data_1<- phds %>%
  group_by(broad_field, year) %>%
  summarize(total_phds= sum(n_phds, na.rm = TRUE))

agg_data_1

#please click on "Zoom" to see in real size, now the sizes are a bit messed up with the small window. 
ggplot(agg_data_1, aes(x= year, y= total_phds, color=broad_field))+
  geom_line(size= 1.2)+
  theme_minimal()+
  labs(title= "Number of awarder Ph.D-s in the US by year",
       x=NULL,
       y= NULL,
       color= "Broad field")+
  scale_color_brewer(palette="Dark2")+
  scale_x_continuous(breaks= pretty_breaks())+
  scale_y_continuous(labels= comma_format())+
  theme(plot.title=element_text(hjust=0.5, 
                                size=20))
  
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
data_task3 <- tt_load("2019-11-05")
names(data_task3)
commute <- data_task3$commute

View(commute)



agg_data_2 <- commute %>%
  group_by(state_region, state_abb) %>%  # Group by state
  summarise(
    total_bike = sum(n[mode == "Bike"]),  # Sum the number of people biking (where mode is "Bike")
    total_walk = sum(n[mode == "Walk"])   # Sum the number of people walking (where mode is "Walk")
  )

View(agg_data_2)  # View the resulting summarized data


ggplot(agg_data_2, aes(x=total_walk,
                      y=total_bike,
                     label= state_abb))+
  geom_point(size = 2, aes(color = state_region)) +
  geom_text(vjust = -1, hjust = 0.5)+
  scale_x_log10(labels = scales::comma_format(big.mark = "."),limits = c(3000, 300000))+
  scale_y_log10(labels = scales::comma_format(big.mark = "."),limits = c(100,100000))+
  theme_light()+
  labs(title="Title number if people walking vs. biking to work in each USA state",
       x= "Number of ppl walking to work (log N)",
       y= "Number of ppl biking to work (log N)",
       color = "State region")+
  
 theme(plot.title=element_text(hjust=0.5, 
                                size=20),
       axis.title= element_text(size=15))
 



```
