---
title: "Assignment 4: Logistic regression"
author: "Marton Kovacs"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(haven)
library(psych)
library(naniar)
library(mice)
library(VIM)
library(misty)
library(corrplot)
library(caret)
library(car)
library(lmtest)
```

# Background story

In this lab assignment you are going to work with data related to the survival of passengers of the RMS Titanic. “The sinking of the Titanic is one of the most infamous shipwrecks in history. On April 15, 1912, during her maiden voyage, the widely considered “unsinkable” RMS Titanic sank after colliding with an iceberg. Unfortunately, there weren’t enough lifeboats for everyone onboard, resulting in the death of 1502 out of 2224 passengers and crew. While there was some element of luck involved in surviving, it seems some groups of people were more likely to survive than others.” (Quote from the Kaggle Titanic Challenge).

For the sake of this assignment, let’s imagine that you are called as an expert to a court case: Kate, one of the survivors of the Titanic accident is suing her __father, Leonardo, for not accompanying Kate and her mother Sue on the trip__ and this way decreasing their chances of survival. The family planned to move to the US back in 1912. __They bought 3rd class tickets for the three of them for 8 British Pounds each. (They did not get cabins with their 3rd class tickets.)__ The plan was that they embark in Southampton and all of them got on board, but Leonardo got separated from them in the rush of passengers during boarding. Later it turned out that Leonardo deliberately got separated from them and got off the boat before it’s departure, to run away and live with his mistress. __Kate was only 4 at the time, and Sue was 20.__ During the accident __Kate got on one of the last lifeboats and was later rescued, but there was no room for Sue on the lifeboat, and she did not survive the disaster.__

Now 20 years later Kate is suing her father for leaving them on the boat, because she thinks that this eventually led to Sue’s death, as the absence of Leonardo decreased their chances of survival.

You are called in as an expert to this court case. Your task is to present a report about whether the presence of Leonardo statistically could have led to an improved chance of survival.

# Dataset

Use the data file called ‘assignment_4_dataset’, from the 'data/' folder.

This is the training dataset of the Titanic dataset from the Kaggle Titanic Challenge (https://www.kaggle.com/c/titanic/overview), a prediction challenge for people who are just starting to learn about machine learning and other statistical prediction techniques. The following description is available for the dataset:

## Metadata

```{r echo = FALSE, warning = FALSE, message = FALSE}
metadata <- readr::read_tsv(here::here("~/Documents/GitHub/Fall-2024/data/assignment_6_metadata_.tsv"))

metadata %>% 
  knitr::kable()
```

# Task

As usual, start with exploring your dataset. Do descriptive and exploratory analysis including visualization to understand the data and to see what type of data you are dealing with. 

You should build a statistical model with which you can accurately estimate Kate’s and Sue’s chances of survival. First you should fit a statistical model (for example a logistic regression model) on the dataset, calculate the regression equation, and use that equation to compute the survival probability for Kate and Sue separately with and without having Leonardo on board the ship with them.

You can use whichever predictor you would like, but you need to build a model that is at least as accurate so that it can correctly predict the outcome value within the sample with at least 72% accuracy for BOTH those who actually survived and who actually died in the disaster. You need to check this in the Classification table. So it is not enough to have 72% overall correct percentage! In order to be able to reach this prediction accuracy you might have to use some special predictors or to do some feature engineering. A comprehensive exploratory analysis including the visualisation of the relationship of different predictors might help in this. Keep in mind that this is one of the most popular prediction challenges on Kaggle, so there is plenty of discussion and guides on the web about different models and features. If you get stuck, you can look these up to improve your prediction performance.

You do not need to check model assumptions in this assignment (but you can do so if you want to and this might help you improve your prediction performance). 

# What to report

When you have arrived at a satisfactory model describe the final model to the reader so that it is clear how is the model built up, and that based on the description the reader could reproduce your model.

Report about the goodness of fit of the model, whether it is significantly better than the null model (based on the AIC and chi^2 test statistics), and how effective is your model at predicting the outcome (based on McFadden R^2, and the correct prediction percentages in the classification table of the final model). Be sure to report the total correct prediction percentage of the final model and also the correct prediction percentages separately for those who actually died, and those who actually survived.

Also, report the statistics describing the coefficients of the predictors in a table format (for each predictor, this table should include the following: logit regression coefficients, Odds ratios, and 95% confidence intervals for the Odds ratios, Chi^2 test statistics and p values, and AIC values for the reduced models). 

Report which were the most influential predictors in the model, and which were the predictors which did not seem to have unique added value to the model.

Write up the regression equation of the model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Finally, report the predicted probability of survival for Kate and Sue separately with and without having Leonardo on board the ship with them. (So you will have to estimate 4 probabilities in total, two for Kate and two for Sue). It is important that this is in the probability scale (since the jury does not know what logit means and how to interpret it).

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses in light of the court case. Based on your results do you find it likely that the presence of Leonardo (spouse to Sue and parent to Kate) would have improved the survival chances of Sue and Kate? What is the best predictor of survival in the model and how does the presence of a spouse and presence of a parent compare to its influence?

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile.

```{r}
titanic_data <- read_sav("~/Documents/GitHub/Fall-2024/data/assignment_6_dataset_.sav")
class(titanic_data)
#format the data for str (however it's just a personal preference)
titanic_data <- zap_formats(titanic_data)
titanic_data <- zap_labels(titanic_data)
titanic_data <- zap_widths(titanic_data)

titanic_data <- as.data.frame(titanic_data)
View(titanic_data)
```

## EDA

```{r}
str(titanic_data)
table(titanic_data$Survived) #How many people survived
table(titanic_data$Pclass) #Most of the people were travelling on third class. 
table(titanic_data$Sex) #Gender distribution
describe(titanic_data$Age) #Descriptive stats for Age
table(titanic_data$SibSp) #Were the passengers with other people from the family?
table(titanic_data$Parch) #Children were alone?
describe(titanic_data$Fare)

#creating binary data for Cabin variable: If there is a cabin number, than 1 if not than 0
titanic_data <- titanic_data %>%
  mutate(Cabin_b = ifelse(Cabin == "", 0, 1),
         Embarked_coded = ifelse(Embarked == "C", 1, ifelse( Embarked == "Q", 2, 3)),
         Sex_coded = ifelse(Sex == "female", 1, 0))

str(titanic_data)
#Verify if it was successful, by randomly:
View(titanic_data)

table(titanic_data$Cabin_b) #Most of the passenger did not have a cabin
table(titanic_data$Embarked) #Where did they step on board

#Is there any missing data? 
colSums(is.na(titanic_data)) #<- Yes at Age there is 177 missing values. 
```

## Clean the data

```{r}
#I created a new variable, because if a problem occurs, it is easier to re-run the modification. 
titanic_data_cleaned <- titanic_data %>%
  select(-PassengerId,
         -Ticket,
         -Cabin,
         -Embarked,
         -Sex)

#Checking whether there is any pattern in missingness, or it is MCAR (Missing Complitely at Random)
mcar_test(titanic_data_cleaned)
vis_miss(titanic_data)
mcar_test(titanic_data %>% select(Age)) #Significant
md.pattern(titanic_data)

 


#Please Step to the next session: After created datatable for Leonardo, and removed the name column, come back here. Because: We needed to assign Leonardo to the original data, to impute his age, and then remove from the OG dataset with the name: with that we are sure no one else is removed.
##################################################################################################################



correlation_matrix <- cor(training_dataset)
corrplot(correlation_matrix, method= "color")
# First we must calculate the outlier variables:
model_1<- glm(Survived ~ Pclass + Sex_coded+ Age+ Parch + Fare+ Cabin_b+ Embarked_coded+ Pclass:Age+  Pclass:Fare + Pclass:Cabin_b+ SibSp:Parch + Cabin_b:Fare,  data= training_dataset, family = "binomial")
summary(model_1)

#Cook's distance
cooks_d <- cooks.distance(model_1)

#Influential points
influential<- which(cooks_d > 4/nrow(training_dataset))

# Create new dataset without outliers
training_final<- training_dataset[-influential, ]



#Assigning interaction between the variables which correlation is min |3|
model_1_1<- glm(Survived ~ Pclass + Sex_coded+ Age+ Parch + Fare+ Cabin_b+ Embarked_coded+ Pclass:Age+  Pclass:Fare + Pclass:Cabin_b+ SibSp:Parch + Cabin_b:Fare,  data= training_final, family = "binomial")
summary(model_1_1)

################################################################################
#Please go to the line 245
```

## Creating a datatable for Sue, Kate, and Leonardo

```{r}

#Trying to find Kate
Kate_precise<- titanic_data %>% 
  filter(grepl("Kate", Name, ignore.case = TRUE) & 
         Embarked == "S" & 
         Pclass == 3 & 
         Cabin_b == 0)             #<- Although it returned results, they are not matching exactly the other given attributes (neither does for Sue and Leonardo)                

Kate <- data.frame(Survived = 1,
                    Pclass = 3,
                    Age = 4,
                    SibSp = 0,
                    Parch = 1,
                    Fare = 8,
                    Cabin_b = 0,
                    Embarked_coded = 3,
                    Sex_coded = 1)


Sue <- data.frame(Survived = 0,
                    Pclass = 3,
                    Age = 20,
                    SibSp = 1,
                    Parch = 1,
                    Fare = 8,
                    Cabin_b = 0,
                    Embarked_coded = 3,
                    Sex_coded = 1)


Leonardo<- data.frame(Survived = NA,
                    Pclass = 3,
                    Name= "Leonardo",
                    Age = NA,
                    SibSp = 1,
                    Parch = 1,
                    Fare = 8,
                    Cabin_b = 0,
                    Embarked_coded = 3,
                    Sex_coded = 0)



#Because Leonardo's Age is unknown I am attaching him to the OG data.
data_with_Leonardo <- rbind(titanic_data_cleaned, Leonardo)
View(data_with_Leonardo)
#Using Mice to inmmpute missing variables (included Leonardo's age)
set.seed(123)
imputed_data <- mice(data_with_Leonardo, meth= "rf", rf.method= "randomForest", m=5, maxit= 10, ntree=10)
#final_list (one method)
complete_with_Leonardo <- complete(imputed_data, action = 1)
View(complete_with_Leonardo)
#Removing Leonardo from the training (OG) dataset.

training_dataset <- complete_with_Leonardo[complete_with_Leonardo$Name != "Leonardo", ]
Leonardo_final <- complete_with_Leonardo[complete_with_Leonardo$Name == "Leonardo", ]
View(training_dataset)
View(Leonardo_final)

Leonardo_final <- Leonardo_final %>%
  select(-Name)

training_dataset <- training_dataset %>%
  select(-Name)
#Please go back to the previous chunk's marked section


################################################################################
#Continue here after the end of the last section
Sue_with_Leonardo <- rbind(Sue, Leonardo_final)

Kate_with_Leonardo <- rbind(Kate, Leonardo_final)



```

## Building the null model + Check the assumptions
The assumtion check was copied here, because I used them to the model building. Sorry for the inconvenience.

```{r}
#Again the model
model_1_1<- glm(Survived ~ Pclass + Sex_coded+ Age+ Parch + Fare+ Cabin_b+ Embarked_coded+ Pclass:Age+  Pclass:Fare + Pclass:Cabin_b+ SibSp:Parch + Cabin_b:Fare,  data= training_final, family = "binomial")
summary(model_1_1)

vif(model_1_1) #<- removing the variables which vif vale is greater than 10
model_1_2<- glm(Survived ~ Pclass + Sex_coded+ Age+ Parch +  Embarked_coded+ SibSp:Parch ,  data= training_final, family = "binomial")
summary(model_1_2)
bptest(model_1_2) #Heteroscedasticity is not present. 
raintest(model_1_2) #Rainbow test -> the model is not biased. 

train_control <- trainControl(method = "cv", number = 10)
AIC(model_1_2)



# Train a logistic regression model with cross-validation: removing predictors based on model_1_1, leaving only the significant ones. 
training_final$Survived <- as.factor(training_final$Survived)
model_cv <- train(Survived ~ Sex_coded+ Age+  Parch +  Embarked_coded+ SibSp:Parch ,  data= training_final, family = "binomial", trControl = train_control)
```

## Building the model

```{r}

#Prediction only for  Kate
prediction_Kate <- predict(model_cv, newdata = Kate, type = "prob")[,2]
print(prediction_Kate)
#Prediction only for Sue
prediction_Sue <- predict(model_cv, newdata = Sue, type = "prob")[,2]
print(prediction_Sue)

# Create combined data frames for scenarios where Leonardo is present with Sue
#New dataset called Sue_with_Leonardo
Sue_with_Leonardo <- Sue
Sue_with_Leonardo$SibSp <- Sue_with_Leonardo$SibSp + Leonardo$SibSp
Sue_with_Leonardo$Pclass <- Sue_with_Leonardo$Pclass + Leonardo$Pclass
prediction_Sue_with_Leonardo <- predict(model_cv, newdata = Sue_with_Leonardo, type = "prob")[,2]
print(prediction_Sue_with_Leonardo)

# Create combined data frames for scenarios where Leonardo is present with Kate
#New dataset called Sue_with_Leonardo
Kate_with_Leonardo <- Kate
Kate_with_Leonardo$SibSp <- Kate_with_Leonardo$SibSp + Leonardo$SibSp
Kate_with_Leonardo$Pclass <- Kate_with_Leonardo$Pclass + Leonardo$Pclass
prediction_Kate_with_Leonardo <- predict(model_cv, newdata = Kate_with_Leonardo, type = "prob")[,2]

print(prediction_Kate_with_Leonardo)
```

# Compare the models

```{r}
comparison_Kate_with_Leonardo <- data.frame(
  Kate = prediction_Kate,
  Kate_with_Leonardo = prediction_Kate_with_Leonardo)

print(comparison_Kate_with_Leonardo) #<- indicating Kate's chance  emerging when Leonardo is with her

comparison_Sue_with_Leonardo <- data.frame(
  Sue = prediction_Sue,
  Sue_with_Leonardo = prediction_Sue_with_Leonardo)

print(comparison_Sue_with_Leonardo). #<- indicating Sue's chance for survival  emerging when Leonardo is with her
```

# Calculate odds ratio and confidence interval-for the variables in the prediction model

```{r}
#Function was used one of our reserach (e.g. Németh et al., 2024) 
CIbinm2<-function(binglm){
  CI<-sapply(c(-1,0,1),function(x) round(exp(binglm$coefficients+x*qnorm(0.975)*sqrt(diag(vcov(binglm)))),3))
  CI<-matrix(CI,ncol=3)
  rownames(CI) <- names(binglm$coefficients)
  colnames(CI) <- c("lower limit","estimate","upper limit"); print(CI)
}
#The base model
CIbinm2(model_1_2)
```

# Calculating if people with Spuses and  parents are more likely tio survive or not
```{r}
model_baseline <-  glm(Survived~ 1, data= training_dataset, family= "binomial")
model_Sib <-  glm(Survived~  SibSp,
                     data=training_dataset, family= "binomial")
model_Parch <-  glm(Survived~  Parch,
                     data=training_dataset, family= "binomial")

anova(model_baseline, model_Sib, test= "LRT")
anova(model_baseline, model_Parch, test= "LRT")
summary(model_Sib)
summary(model_Parch)
CIbinm2(model_Sib)
CIbinm2(model_Parch)
```


# Report the results

```{r}
# If we are inspecting wether the presence of a spouse and a parent is predicting the survival only teh aprent is a significant positive predictor, spouse is not. 
#However if we are deploying a Machine Learning algorithm, and inspecting EXACTLY the case of Kate's and Sue's chances we got the following results: 
#Kate's prediction was correct: if her Father had been on board, her and her mother's chance would emerged, although her chance would decrease a bit. Also important before I set the seed, both chances emerged, indicating an important factor: Age, which was missing at him, so if I would be a court expert I would ask for his age for a correct prediction. 


#The equation for the Logistic regression model which was the base of the ML:
coef(model_1_2)
# model_1_2 = 4.22 + (-1.58 * Pclass) + (3.34 *Sex_coded) + (-0.06* Age) + (0.46 * Parch)+ (-0.25) * Embarked_coded) + (-0.71* Parch* SibSp)

```

#Cited litearture 
Németh, T., Konok, V., & Pogány, Á. (2024). The effect of the COVID-19 lockodwn on children’s screen time and socio-cognitive skills. 15th Dubrovnik Conference on Cognitive Science, 51–52.
 

